Versioned data: why it is needed and how it can be achieved (easily and cheaply)

### 2017-10-17 ms.tex
reworking the discussion a bit

%work on this paragraph more
We acknowledge that the issues we have identified in this article will not come as news to most readers: datasets are constantly improving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact. In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought about versioning and developed specific solutions. But so far these as we mentioned above, these datasets tend to be those managed by large, well-funded institutions or consortia and involve standard data types (e.g., sequences). But for an individual research group working with a specialized type of data, creating their own \texttt{GenBank} is clearly a non-starter. As a result, we argue that there is a substantial need for a easy, cheap, and flexible solution and suggest that adopting a \textsc{vdd} system may be one way forward.

% Moved from above, need to integrate better
We have created \texttt{datastorr} to make \textsc{vdd} easy to implement and manage for \texttt{R} users. And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
% Reword next senetnce
But the design choices we made in \texttt{datastorr} represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with \texttt{GitHub} (or an entirely different service and technology) and accessing versioned data programmatically.  The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.

Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many platforms for storing and distributing versioned controlled data with little or no cost.  The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories. In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples). Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}. Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.  The proliferation of cloud-based tools and open-source software have now reduced these roadblocks considerably.  We argue that by linking together the tools described above, researchers can now produce, curate, and distribute Versioned Data much more easily and cheaply than in the past.

@@

Additionally, one of the greatest benefits of using a cloud-based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}. Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved. As a result adopting a \textsc{vdd} system has the added benefit of facilitating collaboration among research groups.

@@

% subsection only temporary as not allowed by journal guidelines.
% trick is to say something about how things are getting better rapidly, but not scare people away from doing this now.  
\subsection{Future directions}

Technology for data management and distribution is moving forward rapidly.  The workflow we describe above works well with current tools and should work even more quickly with its future replacement.  Here, we highlight a few rapidly developing future tools that researchers should be aware of.

Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.
% Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future

Another area requiring attention is properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data. Likewise, we are optimistic that our funding bodies, research institutions, and scientific colleagues will further recognize the value in adding to, improving, and curating existing datasets in addition to collecting new ones.

>>>>

%work on this paragraph more
We strongly suspect that the issues we have identified in this article will not come as news to most readers: datasets are constantly improving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact. In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought about versioning and developed specific solutions. But so far these as we mentioned above, these datasets tend to be those managed by large, well-funded institutions or consortia and involve standard data types (e.g., sequences). But for an individual research group working with a specialized type of data, creating their own \texttt{GenBank} is clearly a non-starter. As a result, we argue that there is a substantial need for a easy, cheap, and flexible solution and suggest that adopting a \textsc{vdd} system may be one way forward.

% Moved from above, need to integrate better
We have created \texttt{datastorr} to make such a \textsc{vdd} system easy to implement and manage for \texttt{R} users. And as we state above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
% Reword next senetnce
But the design choices we made in \texttt{datastorr} represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with \texttt{GitHub} --- or an entirely different service and technology --- and accessing versioned data programmatically.  The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.

%Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many platforms for storing and distributing versioned controlled data with little or no cost.  The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories. In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples). Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}.

 A central feature of this concept is that the versioned data is maintained in the cloud. This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.  The proliferation of cloud-based tools and open-source software have now reduced these roadblocks considerably.  We argue that by linking together the tools described above, researchers can now produce, curate, and distribute Versioned Data much more easily and cheaply than in the past.  Additionally, one of the greatest benefits of using a cloud-based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}. Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved. As a result adopting a \textsc{vdd} system has the added benefit of facilitating collaboration among research groups.

 While our first attempt \textsc{vdd} system, \texttt{datastorr} is currently being used, there remain several challenges that we anticipate will be overcome as both technology and scientific norms change. For example, because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes. There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}. Another area requiring attention is properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data. Likewise, we are optimistic that our funding bodies, research institutions, and scientific colleagues will further recognize the value in adding to, improving, and curating existing datasets in addition to collecting new ones.

% Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future

@@

@@

% subsection only temporary as not allowed by journal guidelines.
% trick is to say something about how things are getting better rapidly, but not scare people away from doing this now.  
%\subsection{Future directions}

%Technology for data management and distribution is moving forward rapidly.  The workflow we describe above works well with current tools and should work even more quickly with its future replacement.  Here, we highlight a few rapidly developing future tools that researchers should be aware of.

%Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
%There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.
% Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future

%Another area requiring attention is properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data. Likewise, we are optimistic that our funding bodies, research institutions, and scientific colleagues will further recognize the value in adding to, improving, and curating existing datasets in addition to collecting new ones.
